Cell Decompositions for Rank Two Quiver Grassmannians

### 2018-03-06 Grassmannian_cells.tex
Sketch of bijection proof, more details needed

For $m\ge2$ and $1\le k\le n-1$, write $D_m^{(k)}$ for the maximal Dyck path obtained from $D_m$ by removing the first $k$ copies of $D_{m-1}$.
Extending this notation we also set $D_m^{(0)}:=D_m$.
\begin{remark}
  Note that by Theorem~\ref{th:dyck path recursion}, we have a natural identification of $D_m^{(n-1)}$, $m\ge3$, with $D_{m-1}^{(1)}$.
\end{remark}
For fixed $m\ge2$ and $1\le r\le n-1$, we write $D_{m-1,r}$ for the $r$-th copy of $D_{m-1}$ inside $D_m$.
For notational convenience we also set $D_{m-1,n}:=D_m^{(n-1)}$ even though this maximal Dyck path does not identify with a copy of $D_{m-1}$.
For $m\ge2$ and $1\le k\le n-1$, the maximal Dyck paths $D_{m-1,r}$, $k+1\le r\le n$, naturally identify with subpaths of $D_m^{(k)}$.

We identify the edges of $D_m$ with the ordered set $E_m=\{1,\ldots,u_m+u_{m-1}\}$, where edges of $D_m$ are taken in the natural order beginning from $(0,0)$.
Let $E_m=H_m\sqcup V_m$, where $H_m=\{h_1,\ldots,h_{u_m}\}$ and $V_m=\{v_1,\ldots,v_{u_{m-1}}\}$ denote the horizontal and vertical edges of $D_m$ respectively.
Following Theorem~\ref{th:dyck path recursion}, we partition the edges as $E_m=\bigsqcup_{r=1}^n E_{m-1,r}$, where $E_{m-1,r}$ denotes the edges of $D_{m-1,r}$.
The set $E_{m-1,r}$ is naturally partitioned into its subsets $H_{m-1,r}$ and $V_{m-1,r}$ of horizontal and vertical edges.

Given edges $e,e'\in E_m$ with $e<e'$, write $ee'$ for the shortest subpath of $D_m$ containing $e$ and $e'$, in particular $ee$ is the subpath containing the single edge $e$.
\begin{definition}
  \label{def:compatibility}
  A pair of subsets $S_H\subset H_m$ and $S_V\subset V_m$ is called \emph{compatible} if: 
  for each pair $(h,v)\in S_H\times S_V$ with $h<v$ there exists an edge $e\in hv$ so that at least one of the following holds
  \begin{equation}
    \label{eq:hgc}
    e\ne v\qquad\text{and}\qquad |he\cap V_m|=n|he\cap S_H|
  \end{equation}
  or
  \begin{equation}
    \label{eq:vgc}
    e\ne h\qquad\text{and}\qquad |ev\cap H_m|=n|ev\cap S_V|.
  \end{equation}
  Write $\cC_m$ for the collection of all pairs $(S_H,S_V)$ which are compatible as above.
\end{definition}
\begin{remark}
  This notion of compatibility extends naturally to the maximal Dyck paths $D_m^{(k)}$, $1\le k\le n-1$.  Write $\cC_m^{(k)}$ for the set of all compatible pairs in $D_m^{(k)}$.
\end{remark}

The recursive structure of the maximal Dyck paths from Theorem~\ref{th:dyck path recursion} gives rise to a recursive characterization of compatible pairs.
\begin{definition}
  \cite[Definition 3.11]{rupel}
  \label{def:piecewise compatibility}
  A pair of subsets $S_H\subset H_m$ and $S_V\subset V_m$ is called \emph{piecewise compatible} if, for each $1\le r\le n$, one of the conditions \eqref{eq:hgc} or \eqref{eq:vgc} is satisfied for each pair $(h,v)\in S_H\times S_V$ with $h\in H_{m-1,r}$ and $v\in V_{m-1,r}$.
\end{definition}
\begin{remark}
  The notion of piecewise compatibility naturally extends to the maximal Dyck paths $D_m^{(k)}$, $1\le k\le n-1$.
\end{remark}

@@

\begin{corollary}
  For $m\ge4$ and $0\le k\le n-1$, consider $S_H\subset H_m^{(k)}$ and $S_V\subset V_m^{(k)}$ so that $(S_H,S_V)$ is piecewise compatible.
  Assume $\big(S_H^{(k+1)},S_V^{(k+1)}\big)\in\cC_m^{(k+1)}$.
  Then $(S_H,S_V)$ is not compatible if and only if $H_{m-2}^{(k+1)}\subset S_{H,k+1}$ and $V_m^{(k+1)}\subset S_V$.
\end{corollary}
\begin{proof}
  We begin with the reverse implication.
  First note that there are $(n-k)u_{m-1}-u_{m-2}$ horizontal edges and $(n-k)u_{m-2}-u_{m-3}$ vertical edges in $D_m^{(k)}$. 
  It follows that $V_{m-2}^{(k+1)}\sqcup V_m^{(k+1)}$ contains $n(n-k)u_{m-3}-nu_{m-4}$ vertical edges and $H_{m-2}^{(k+1)}\sqcup H_m^{(k+1)}$ contains $n(n-k)u_{m-2}-nu_{m-3}$ horizontal edges.

  Assuming $H_{m-2}^{(k+1)}\subset S_{H,k+1}$ and $S_V^{(k+1)}=V_m^{(k+1)}$, we have $S_V\cap V_{m-2}^{(k+1)}=\varnothing$ and $S_H\cap H_m^{(k+1)}=\varnothing$ by piecewise compatibility.
  Let $h\in H_m^{(k)}$ be the horizontal edge corresponding to the first horizontal edge of $H_{m-2}^{(k+1)}$.
  Then, since there are $(n-k)u_{m-3}-u_{m-4}$ horizontal edges in $H_{m-2}^{(k+1)}$, the local shadow path $D(h;S_H)$ contains $n\big((n-k)u_{m-3}-u_{m-4}\big)$ vertical edges and is thus equal to $hv_{u_{m-1}}$.
  Similarly, the local shadow path $D(v_{u_{m-1}};S_V)$ is also equal to $hv_{u_{m-1}}$.
  In particular, neither of the compatibility conditions of Definition~\ref{def:compatibility} are satisfied for the path $hv_{u_{m-1}}$ and so $(S_H,S_V)$ is not compatible.

  For the forward implication, we work by induction on $m\ge4$.
  Consider a pair $(S_H,S_V)$ for $D_4^{(k)}$ as above which is not compatible.
  Following Theorem~\ref{th:blocking edge conditions}, write $h\in H_4^{(k)}$ for the blocking edge of $S_H$.
  Then the number of vertical edges in the local shadow path $D(h;S_H)=hv_{u_3}$ must be divisible by $n$. 
  Since $(S_H^{(k+1)},S_V^{(k+1)})$ is compatible, we must have $h\in H_{3,k}$.
  But observe that $|V_m^{(k)}|=(n-k)n-1$ and so the divisibility condition above implies $h\in H_{3,k}^{(n-1)}$.
  But $S_V$ is strongly right-justified with respect to $h$ and thus the number of horizontal edges in $D(v_{u_3};S_V)=hv_{u_3}$ is divisible by $n$.
  Identifying $H_{3,k}^{(n-1)}$ with $H_2^{(1)}$, this divisibility condition only occurs when $h$ is the first horizontal edge in $H_2^{(k+1)}\subset H_2^{(1)}$.
  Then by piecewise compatibility, the vertical edge of $H_2^{(1)}$ cannot be an element of $S_V$ and we must have $V_4^{(k+1)}\subset S_V$.
  By piecewise compatibility again, this implies $H_4^{(k+1)}\cap S_H=\varnothing$ and so $D(h;S_H)=hv_{u_3}$ implies $H_2^{(k+1)}\subset S_H$.

  To continue, let $(S_H,S_V)$ be a pair for $D_m$, $m\ge5$, which is not compatible.
  Write $\varphi:H_{m-1}\to V_m$ for the bijection given by $\varphi(h_i)=v_i$ for $1\le i\le u_{m-1}$.
  For any subset $T\subset H_{m-1}$, set $\varphi^*(T)=V_m\setminus\varphi(T)$.
  Clearly, the map $\varphi^*$ gives a bijection between subsets of $H_{m-1}$ and subsets of $V_m$.
  In Section 3.2 of \cite{rupel}, a new pair of subsets $\big((\varphi^*)^{-1}S_V,\Omega^{-1}S_H\big)$ for $D_{m-1}$ is given, we refer the reder to \emph{loc. cit} for notation.
  By \cite[Proposition 3.10]{rupel}, the pair $\big((\varphi^*)^{-1}S_V,\Omega^{-1}S_H\big)$ is not compatible, but is piecewise compatible by \cite[proposition 3.16]{rupel}.
  Thus by induction, we must have $H_{m-3}^{(k+1)}\subset(\varphi^*)^{-1}S_V$ and $V_{m-1}^{(k+1)}\subset\Omega^{-1}S_H$.
  It follows from piecewise compatibility that $H_{m-1}^{(k+1)}\cap(\varphi^*)^{-1}S_V=\varnothing$.
  But then by the definition of $\varphi^*$ we have $V_{m-2}^{(k+1)}\cap S_V=\varnothing$ and $S_V^{(k+1)}=V_m^{(k+1)}$ so that $D(v_{u_{m-1}};S_V)=hv_{u_{m-1}}$ with $h$ as in the first case above.
  Since $(S_H,S_V)$ is not compatible, Theorem~\ref{th:blocking edge conditions} states that $h$ must be the blocking edge for $S_H$ and we must have $D(h;S_H)=hv_{u_{m-1}}$.
  But this can only occur if $H_{m-2}^{(k+1)}\subset S_H$ since $H_m^{(k+1)}\cap S_H=\varnothing$ by piecewise compatibility.
\end{proof}

The following result is an immediate consequence of the combinatorial construction of rank 2 cluster variables \cite{llz} and the categorification of these variables using representations of $K(n)$ \cite{cc,ck}.
\begin{theorem}\cite{llz}
  For each $m\ge1$ and $\bfe\in\ZZ_{\ge0}^2$, we have $\chi\big(\Gr_\bfe(P_m)\big)=\big|\{(S_H,S_V)\in\cC_m:|S_H|=u_m-e_1,|S_V|=e_2\}\big|$.
\end{theorem}

Our goal is to show that the compatible pairs provide a natural labeling for the cells of $\chi\big(\Gr^Q_\bfe(P_m)\big)$ found in Theorem~ ?%\ref{???}.

>>>>

For $m\ge2$ and $1\le r\le n-1$, write $D_m^{[r]}$ for the maximal Dyck path obtained from $D_m$ by removing the first $r$ copies of $D_{m-1}$.
Extending this notation we also set $D_m^{[0]}:=D_m$.
\begin{remark}
  Note that by Theorem~\ref{th:dyck path recursion}, we have a natural identification of $D_m^{[n-1]}$, $m\ge3$, with $D_{m-1}^{[1]}$.
\end{remark}
For fixed $m\ge2$ and $1\le i\le n-1$, we write $D_{m-1,i}$ for the $i$-th copy of $D_{m-1}$ inside $D_m$.
For notational convenience we also set $D_{m-1,n}:=D_m^{[n-1]}$ even though this maximal Dyck path does not identify with a copy of $D_{m-1}$.
For $m\ge2$ and $1\le r\le n-1$, the maximal Dyck paths $D_{m-1,i}$, $r+1\le i\le n$, naturally identify with subpaths of $D_m^{[r]}$.

We identify the edges of $D_m$ with the ordered set $E_m=\{1,\ldots,u_m+u_{m-1}\}$, where edges of $D_m$ are taken in the natural order beginning from $(0,0)$.
Let $E_m=H_m\sqcup V_m$, where $H_m=\{h_1,\ldots,h_{u_m}\}$ and $V_m=\{v_1,\ldots,v_{u_{m-1}}\}$ denote the horizontal and vertical edges of $D_m$ respectively.
Following Theorem~\ref{th:dyck path recursion}, we partition the edges as $E_m=\bigsqcup_{i=1}^n E_{m-1,i}$, where $E_{m-1,i}$ denotes the edges of $D_{m-1,i}$.
The set $E_{m-1,i}$ is naturally partitioned into its subsets $H_{m-1,i}$ and $V_{m-1,i}$ of horizontal and vertical edges.

Given edges $e,e'\in E_m$ with $e<e'$, write $ee'$ for the shortest subpath of $D_m$ containing $e$ and $e'$, in particular $ee$ is the subpath containing the single edge $e$.
\begin{definition}
  \label{def:compatibility}
  A pair of subsets $S_H\subset H_m$ and $S_V\subset V_m$ is called \emph{compatible} if: 
  for each pair $(h,v)\in S_H\times S_V$ with $h<v$ there exists an edge $e\in hv$ so that at least one of the following holds
  \begin{equation}
    \label{eq:hgc}
    e\ne v\qquad\text{and}\qquad |he\cap V_m|=n|he\cap S_H|
  \end{equation}
  or
  \begin{equation}
    \label{eq:vgc}
    e\ne h\qquad\text{and}\qquad |ev\cap H_m|=n|ev\cap S_V|.
  \end{equation}
  Write $\cC_m$ for the collection of all pairs $(S_H,S_V)$ which are compatible as above.
\end{definition}
\begin{remark}
  This notion of compatibility extends naturally to the maximal Dyck paths $D_m^{[r]}$, $1\le r\le n-1$.  Write $\cC_m^{[r]}$ for the set of all compatible pairs in $D_m^{[r]}$.
\end{remark}

The recursive structure of the maximal Dyck paths from Theorem~\ref{th:dyck path recursion} gives rise to a recursive characterization of compatible pairs.
\begin{definition}
  \cite[Definition 3.11]{rupel}
  \label{def:piecewise compatibility}
  A pair of subsets $S_H\subset H_m$ and $S_V\subset V_m$ is called \emph{piecewise compatible} if, for each $1\le r\le n$, one of the conditions \eqref{eq:hgc} or \eqref{eq:vgc} is satisfied for each pair $(h,v)\in S_H\times S_V$ with $h\in H_{m-1,i}$ and $v\in V_{m-1,i}$.
\end{definition}
\begin{remark}
  The notion of piecewise compatibility naturally extends to the maximal Dyck paths $D_m^{[r]}$, $1\le r\le n-1$.
\end{remark}

@@

\begin{corollary}
  \label{cor:piecewise not compatible}
  For $m\ge4$ and $0\le r\le n-1$, consider $S_H\subset H_m^{[r]}$ and $S_V\subset V_m^{[r]}$ so that $(S_H,S_V)$ is piecewise compatible.
  Assume $\big(S_H^{[r+1]},S_V^{[r+1]}\big)\in\cC_m^{[r+1]}$.
  Then $(S_H,S_V)$ is not compatible if and only if $H_{m-2}^{[r+1]}\subset S_{H,r+1}$ and $V_m^{[r+1]}\subset S_V$.
\end{corollary}
\begin{proof}
  We begin with the reverse implication.
  First note that there are $(n-r)u_{m-1}-u_{m-2}$ horizontal edges and $(n-r)u_{m-2}-u_{m-3}$ vertical edges in $D_m^{[r]}$. 
  It follows that $V_{m-2}^{[r+1]}\sqcup V_m^{[r+1]}$ contains $n(n-r)u_{m-3}-nu_{m-4}$ vertical edges and $H_{m-2}^{[r+1]}\sqcup H_m^{[r+1]}$ contains $n(n-r)u_{m-2}-nu_{m-3}$ horizontal edges.

  Assuming $H_{m-2}^{[r+1]}\subset S_{H,r+1}$ and $S_V^{[r+1]}=V_m^{[r+1]}$, we have $S_V\cap V_{m-2}^{[r+1]}=\varnothing$ and $S_H\cap H_m^{[r+1]}=\varnothing$ by piecewise compatibility.
  Let $h\in H_m^{[r]}$ be the horizontal edge corresponding to the first horizontal edge of $H_{m-2}^{[r+1]}$.
  Then, since there are $(n-r)u_{m-3}-u_{m-4}$ horizontal edges in $H_{m-2}^{[r+1]}$, the local shadow path $D(h;S_H)$ contains $n\big((n-r)u_{m-3}-u_{m-4}\big)$ vertical edges and is thus equal to $hv_{u_{m-1}}$.
  Similarly, the local shadow path $D(v_{u_{m-1}};S_V)$ is also equal to $hv_{u_{m-1}}$.
  In particular, neither of the compatibility conditions of Definition~\ref{def:compatibility} are satisfied for the path $hv_{u_{m-1}}$ and so $(S_H,S_V)$ is not compatible.

  For the forward implication, we work by induction on $m\ge4$.
  Consider a pair $(S_H,S_V)$ for $D_4^{[r]}$ as above which is not compatible.
  Following Theorem~\ref{th:blocking edge conditions}, write $h\in H_4^{[r]}$ for the blocking edge of $S_H$.
  Then the number of vertical edges in the local shadow path $D(h;S_H)=hv_{u_3}$ must be divisible by $n$. 
  Since $\big(S_H^{[r+1]},S_V^{[r+1]}\big)$ is compatible, we must have $h\in H_{3,r}$.
  But observe that $|V_m^{[r]}|=(n-r)n-1$ and so the divisibility condition above implies $h\in H_{3,r}^{[n-1]}$.
  But $S_V$ is strongly right-justified with respect to $h$ and thus the number of horizontal edges in $D(v_{u_3};S_V)=hv_{u_3}$ is divisible by $n$.
  Identifying $H_{3,r}^{[n-1]}$ with $H_2^{[1]}$, this divisibility condition only occurs when $h$ is the first horizontal edge in $H_2^{[r+1]}\subset H_2^{[1]}$.
  Then by piecewise compatibility, the vertical edge of $H_2^{[1]}$ cannot be an element of $S_V$ and we must have $V_4^{[r+1]}\subset S_V$.
  By piecewise compatibility again, this implies $H_4^{[r+1]}\cap S_H=\varnothing$ and so $D(h;S_H)=hv_{u_3}$ implies $H_2^{[r+1]}\subset S_H$.

  To continue, let $(S_H,S_V)$ be a pair for $D_m$, $m\ge5$, which is not compatible.
  Write $\varphi:H_{m-1}\to V_m$ for the bijection given by $\varphi(h_i)=v_i$ for $1\le i\le u_{m-1}$.
  For any subset $T\subset H_{m-1}$, set $\varphi^*(T)=V_m\setminus\varphi(T)$.
  Clearly, the map $\varphi^*$ gives a bijection between subsets of $H_{m-1}$ and subsets of $V_m$.
  In Section 3.2 of \cite{rupel}, a new pair of subsets $\big((\varphi^*)^{-1}S_V,\Omega^{-1}S_H\big)$ for $D_{m-1}$ is given, we refer the reder to \emph{loc. cit} for notation.
  By \cite[Proposition 3.10]{rupel}, the pair $\big((\varphi^*)^{-1}S_V,\Omega^{-1}S_H\big)$ is not compatible, but is piecewise compatible by \cite[proposition 3.16]{rupel}.
  Thus by induction, we must have $H_{m-3}^{[r+1]}\subset(\varphi^*)^{-1}S_V$ and $V_{m-1}^{[r+1]}\subset\Omega^{-1}S_H$.
  It follows from piecewise compatibility that $H_{m-1}^{[r+1]}\cap(\varphi^*)^{-1}S_V=\varnothing$.
  But then by the definition of $\varphi^*$ we have $V_{m-2}^{[r+1]}\cap S_V=\varnothing$ and $S_V^{[r+1]}=V_m^{[r+1]}$ so that $D(v_{u_{m-1}};S_V)=hv_{u_{m-1}}$ with $h$ as in the first case above.
  Since $(S_H,S_V)$ is not compatible, Theorem~\ref{th:blocking edge conditions} states that $h$ must be the blocking edge for $S_H$ and we must have $D(h;S_H)=hv_{u_{m-1}}$.
  But this can only occur if $H_{m-2}^{[r+1]}\subset S_H$ since $H_m^{[r+1]}\cap S_H=\varnothing$ by piecewise compatibility.
\end{proof}

The following result is an immediate consequence of the combinatorial construction of rank 2 cluster variables \cite{llz} and the categorification of these variables using representations of $K(n)$ \cite{cc,ck}.
\begin{theorem}
  \label{th:combinatorial euler}
  \cite{llz}
  For each $m\ge1$ and $\bfe\in\ZZ_{\ge0}^2$, we have $\chi\big(\Gr_\bfe(P_m)\big)=\big|\{(S_H,S_V)\in\cC_m:|S_H|=u_m-e_1,|S_V|=e_2\}\big|$.
\end{theorem}

Our goal is to give a geometric explanation for this by showing that the compatible pairs provide a natural labeling for the cells of $\chi\big(\Gr^Q_\bfe(P_m)\big)$ found in Theorem~\ref{celldec}.
We accomplish this by providing a bijection between the compatible pairs as in Theorem~\ref{th:combinatorial euler} and the successor closed sets of vertices in the 2-quivers $\cQ_m^{[r]}$ used in Theorem~\ref{thm:2quivercells} to describe the nonempty cells.
\begin{theorem}
  \label{th:compatible cells}
  Compatible pairs label cells in $\Gr_\bfe(P_m^V)$.
\end{theorem}
\begin{proof}
  The recursive construction of the 2-quivers $\cQ_m^{[r]}$ provides a natural ordering of the vertices in the underlying quiver $Q_m^{[r]}$.
  This provides a bijection of these vertices with the edges of $D_m^{[r]}$ whereby vertices covering the vertex 1 (resp.\ vertex 2) of $K(n)$ correspond to horizontal edges (resp.\ vertical edges).
  Then Corollary~\ref{cor:piecewise not compatible} shows that under this bijection a subset $\beta\in(Q_m^{[r]})_0$ is strong successor closed in $\cQ_m^{[r]}$ if and only if the corresponding pair of subsets $\big(S_H(\beta),S_V(\beta)\big)$ is compatible.
\end{proof}
